---
title: "STAT 201A Lab (11/20)"
author: "Kenneth Chen"
date: '2023-11-26'
output: pdf_document
---
## Problem 1a

$$
P = \begin{bmatrix}
    0.2 & 0.7 & 0.1 \\
    0.2 & 0.5 & 0.3 \\
    0.2 & 0.4 & 0.4
  \end{bmatrix}
$$

## Problem 1b

```{r}
set.seed(24)
P <- matrix(c(0.2, 0.7, 0.1, 0.2, 0.5, 0.3, 0.2, 0.4, 0.4), nrow = 3, byrow = T)
P
simulate_mc <- function(n, P, X0) {
    # Generates sequences of length n (including X0)
    #   P: transition probability matrix
    #   X0: initial state
    cur_X <- X0
    states <- 1:3
    MC_seq <- numeric(n)
    MC_seq[1] <- X0
    for (i in 2:n) {
        transition_probs <- P[cur_X, ]
        next_X <- sample(states, size = 1, prob = transition_probs)
        MC_seq[i] <- next_X
        cur_X <- next_X
    }
    MC_seq
}
plot(1:100, simulate_mc(100, P, 1), 
     xlab = "i", ylab = "State", 
     type = "b", pch = 19,
     main = "Realization of Markov Chain (100 Steps, X0 = 1)")
```

## Problem 2a

```{r}
eigs <- eigen(t(P))
stationary_state <- eigs$vectors[, 1]
stationary_state_norm <- stationary_state / sum(stationary_state)
eigs
```

Solving numerically, we see that `r stationary_state` is the solution to $(P^T - I)\pi_\infty = 0$. Normalized, we have $\pi_\infty$ = `r stationary_state_norm`.

## Problem 2b

We plot the convergence for an initial $\pi_0$ close to $\pi_\infty$:

```{r}
# initial distn close to pi_infinity
close_initial <- c(0.1, 0.6, 0.3)
pis <- list()
pis[[1]] <- close_initial
n <- 100
# calculates L2 norm
l2norm <- function(u, v) {
    sqrt(sum((u - v)^2))
}
l2_norms <- numeric(n)
for (i in 2:n) {
    pis[[i]] <- pis[[i - 1]] %*% P
    l2_norms[i] <- l2norm(pis[[i]], stationary_state_norm)
}

plot(1:(n - 1), l2_norms[2:n], type = "l",
     xlab = "i", ylab = "L2 Norm", 
     main = "Convergence of Probability Distribution (Starting Close)")
```

We plot the convergence for an initial $\pi_0$ far from $\pi_\infty$:

```{r}
# initial distn far from pi_infinity
far_initial <- c(0.7, 0.2, 0.1)
pis <- list()
pis[[1]] <- far_initial
n <- 100
# calculates L2 norm
l2norm <- function(u, v) {
    sqrt(sum((u - v)^2))
}
l2_norms <- numeric(n)
for (i in 2:n) {
    pis[[i]] <- pis[[i - 1]] %*% P
    l2_norms[i] <- l2norm(pis[[i]], stationary_state_norm)
}

plot(1:(n - 1), l2_norms[2:n], type = "l",
     xlab = "i", ylab = "L2 Norm", 
     main = "Convergence of Probability Distribution (Starting Far)")
```

We see $\pi_i$ converges to $\pi_\infty$ ($\pi_i \rightarrow \pi_\infty$) quite quickly regardless of the specific $\pi_0$ value.

## Problem 3a

```{r sims, cache = T}
# Function that returns arrival time
until_X3 <- function(X0) {
    t <- 0
    cur_X <- X0
    while (cur_X != 3) {
        cur_X <- simulate_mc(2, P, cur_X)[2]
        t <- t + 1
    }
    t
}

# Simulate for X0 = 1
N <- 1e5
arrivals_1 <- numeric(N)
for (i in 1:N) {
    arrivals_1[i] <- until_X3(1)
}
hist(arrivals_1, xlab = "Arrival Time for X0 = 1",
     main = "Histogram of Arrival Times for X0 = 1")

# Simulate for X0 = 2
arrivals_2 <- numeric(N)
for (i in 1:N) {
    arrivals_2[i] <- until_X3(2)
}
hist(arrivals_2, xlab = "Arrival Time for X0 = 2",
     main = "Histogram of Arrival Times for X0 = 2")

# mean arrival times
mean_arrival_time1 <- mean(arrivals_1)
mean_arrival_time2 <- mean(arrivals_2)
mean_arrival_time1
mean_arrival_time2
```

We see that starting from $X_0 = 1$, we have a mean arrival time of `r mean_arrival_time1` and starting from $X_0 = 2$, we have a mean arrival time of `r mean_arrival_time2`.
